---
title: "Shrock2020_eda"
output: html_notebook
---

Set working directory to current directory, load necessary packages and scripts 
```{r}
# set working directory to where script is
SCRIPT_PATH <- rstudioapi::getSourceEditorContext()$path
setwd(dirname(SCRIPT_PATH))

# load packages 
library(Peptides)
library(readxl)
library(dplyr)
library(readr)
library(tidyr)
library(tidyverse)
library(tibble)
library(data.table)
library(styler)
library(ggplot2)
source("helper_funcs.R")
```


save "raw" dataframes as RDS objects for faster loading in the future
```{r}
SHROCK_DIR <- "~/hdsi-vector-ai/data/Shrock_2020/"
# pat_meta <- read_csv(file.path(SHROCK_DIR,
#                                "Shrock_2020_patient-metadata.csv"))
# cov_z <- read_csv(file.path(SHROCK_DIR,
#                             "Shrock_2020_Coronavirus-screen-IgG-Zscores.csv"))
# vir_z <- read_csv(file.path(SHROCK_DIR,
#                             "Shrock_2020_VirScan-IgG-Zscores.csv"))
cov_lib <- read_csv(file.path(SHROCK_DIR,
                              "Shrock_2020_Coronavirus-library.csv"))
vir_lib <- read_csv(file.path(SHROCK_DIR,
                              "Shrock_2020_VirScan-library.csv"))
                               

# DF_NAMES <- c("pat_meta", "cov_z", "vir_z", "cov_lib", "vir_lib")
# SAVE_DIR <- "../data/Shrock_2020/rds"
# FORMAT <- "rds"
# 
# for (NAME in DF_NAMES) {
#     df <- get(NAME) 
#     save_as(df, save_path = SAVE_DIR, name = NAME, format = FORMAT)
# }
```

```{r}
# RDS_PATH <- "../data/rds"
# SHROCK_KEY <- c("cov", "vir", "pat")
# FORMAT <- "rds"
# # load all Shrock related dataframes that were previously converted from .csv to .rds (should be quicker)
# df_ls <- load_filtered_data(directory = RDS_PATH, format = FORMAT, keywords = SHROCK_KEY)
```

Format dataframes 
```{r}
# # make column names valid R variables
# df_ls <- lapply(df_ls, function(df)  {
#   colnames(df) <- make.names(gsub(" ", "_", colnames(df)), unique = TRUE)
#   return(df)
# })
# 
# # separate out each dataframe
# cov_lib <- df_ls$cov_lib.rds
# cov_z <- df_ls$cov_z.rds
# pat_meta <- df_ls$pat_meta.rds
# vir_lib <- df_ls$vir_lib.rds
# vir_z <- df_ls$vir_z.rds
# 
# # Actually, vir_lib has two columns named "id" -- one column starts from 0 and the other starts from 1
# to be consistent with the 0-indexing happening in cov_lib, I'll call the column in vir_lib starting with 0
# "id" and the column starting with 1 "id_1" and use "id" for now. 
# # Also, cov_lib contains a column that looks like peptide ids but it is named something else; reassign column name "id"
# colnames(cov_lib)[1] <- "id"
# colnames(vir_lib)[17] <- "id_1"


# DF_NAMES <- c("pat_meta", "cov_z", "vir_z", "cov_lib", "vir_lib")
# SAVE_DIR <- "../data/processed/Shrock_2020_formatted"
# FORMAT <- "rds"
# 
# for (NAME in DF_NAMES) {
#     df <- get(NAME)
#     save_as(df, save_path = SAVE_DIR, name = NAME, format = FORMAT)
# }

```
```{r}
# 241104TODO: divide this notebook into two separate notebooks, one for data formatting and one eda proper


############ EDA starts here ###################


DF_NAMES <- c("pat_meta", "cov_z", "vir_z", "cov_lib", "vir_lib")
FORM_PATH <- "../data/processed/Shrock_2020_formatted"
FORMAT <- "rds"
formatted_df_ls <- load_filtered_data(directory = FORM_PATH, format = FORMAT, keywords = DF_NAMES)
cov_lib <- formatted_df_ls$cov_lib.rds
cov_z <- formatted_df_ls$cov_z.rds
pat_meta <- formatted_df_ls$pat_meta.rds
vir_lib <- formatted_df_ls$vir_lib.rds
vir_z <- formatted_df_ls$vir_z.rds
```


```{r}
PRO_DAT_PATH <- dirname(FORM_PATH) # data/processed
VIZ_PATH <- "../results/eda/viz"
MERGED_NAME <- "pat_meta_vir_z_2024-10-22.rds"
# load merged patient metadata--virscan z score dataframe 
pat_vir_z <- readRDS(file.path(PRO_DAT_PATH, MERGED_NAME)) # is a giant file btw will take a while to load
```


```{r}
pat_vz <- pat_vir_z # reassign name for brevity

# check data type of each column in pat_vz
# returns: 
     #        id      rep_value        z_score COVID19_status   Hospitalized         sample            rep 
     # "numeric"    "character"      "numeric"    "character"    "character"      "integer"    "character" 
# column_classes <- sapply(pat_vir_z, class)
# print(column_classes)

# encode covid status and hospitalized status as factors
pat_vz <- pat_vz %>% mutate(across(c(COVID19_status, Hospitalized), as.factor))
FACTORIZED_NAME <- "pat_meta_vir_z_2024-11-05.rds"
save_as(pat_vz, FACTORIZED_NAME, PRO_DAT_PATH, "rds")

# # test plotting for a small subset
# subset_i <- ceiling((1e-7)*dim(pat_vz)[1]) 
# pat_vz_subset <- pat_vz[1:subset_i, ]
# 
# ggplot(pat_vz_subset, aes(x = z_score)) +
#   geom_histogram(bins=10) +
#   labs(title = "histogram of the first 10 virscan z-scores", x = "z-score", y = "freq") +
#   theme_minimal()

# ggsave(filename = "pat_vz_10zscores_test.png", path = VIZ_PATH)

```


## Virscan z-score summary statistics (as is)

```{r}

pat_vz %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

```






```{r}
# Histogram of all z-scores (original scale)
vz_hist_orig <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "Histogram of All Virscan Z-Scores", x = "Z-Score (bin size = 0.5)", y = "Frequency") +
  xlim(-10, 10) +
  theme_minimal()

# Histogram of all z-scores (log scale)
vz_hist_log <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(aes(y = after_stat(count + 10)), binwidth = 0.5, fill = "gray", color = "black") + 
  labs(title = "Histogram of All Virscan Z-Scores (Log Scale)", x = "Z-Score (bin size = 0.5)", y = "Log(Frequency)") +
  xlim(-10, 10) +
  scale_y_log10() +  
  theme_minimal()

# ggsave(filename = "vz_hist_orig.png", plot = vz_hist_orig, path = VIZ_PATH)
# ggsave(filename = "vz_hist_log.png", plot = vz_hist_log, path = VIZ_PATH)

# Display the plots
vz_hist_orig
vz_hist_log

```


Not only does the data look right tailed but ALSO There are almost a billion z-scores. Maybe want to look at a random sample... (see "filtered rs" cell)
We also need to remove outliers (max z-score is 95081.95)

```{r}
# remove extreme outliers and missing values
# post-processing summary statistics 

# outlier threshold: remove values outside of 3 sds 
mean_z <- mean(pat_vz$z_score, na.rm = TRUE)
sd_z <- sd(pat_vz$z_score, na.rm = TRUE)

Z_CUTOFF <- 3
z_outliers <- pat_vz$z_score[abs(pat_vz$z_score) > Z_CUTOFF * sd_z]
length(z_outliers) # 236087 outliers
head(z_outliers)

# only keep values with z-scores within 3 sds 
pat_vz_3sd <- pat_vz %>% 
  filter(!z_score %in% z_outliers)

# summary statistics post-filtering 
pat_vz_3sd %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE)
  )

```

```{r}

# plots
vz_hist_3sd <- ggplot(pat_vz_3sd, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "histogram of Virscan z-scores, excluding outliers", x = "z-score (bin size 0.5)", y = "frequency") +
  theme_minimal()
ggsave(filename = "vz_hist_3sd.png", plot = vz_hist_3sd, path = VIZ_PATH)

# histogram of all z-scores on log scale
vz_hist_3sd_log <- ggplot(pat_vz_3sd, aes(x = z_score)) +
  geom_histogram(aes(y = after_stat(count + 10)), binwidth = 0.5, fill = "gray", color = "black") + 
  labs(title = "histogram of Virscan z-scores, excluding outliers (Log Scale)", x = "z-score (bin size 0.5)", y = "log(frequency)") +
  scale_y_log10() +  
  theme_minimal()
ggsave(filename = "vz_hist_3sd_log.png", plot = vz_hist_3sd_log, path = VIZ_PATH)

vz_hist_3sd 
vz_hist_3sd_log
```


Now take random sample (say n = 10e6) from filtered z-scores

```{r, filtered-rs}
set.seed(2024)
# pat_vz_3sd_random-sample
SAMPLE_SIZE <- 100000
pat_vz_rs <- pat_vz_3sd %>% sample_n(SAMPLE_SIZE)

pat_vz_rs_summary <- pat_vz_rs %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

vz_hist_3sd_rs10e6 <- ggplot(pat_vz_rs, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +
  labs(title = "histogram of 10^6 samples of Virscan z-scores", x = "z-Score", y = "frequency") +
  theme_minimal()
ggsave(filename = "vz_hist_orig.png", plot = vz_hist_3sd_rs10e6, path = VIZ_PATH)

pat_vz_rs_summary
vz_hist_3sd_rs10e6

```




# Subgroups on covid and hospitalization status 

# TODO write summarize as a function 

```{r}
vz_cpos <- pat_vz_3sd %>% filter(COVID19_status == "positive")
vz_cneg <- pat_vz_3sd %>% filter(!COVID19_status == "positive")

vz_cpos_summary <- vz_cpos %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

vz_cneg_summary <- vz_cneg %>%
   summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

vz_cpos_summary
vz_cneg_summary
```

```{r}
# group by covid+ vs. covid-
# summary statistics 
# histograms
# patient-metadata_virscan-z-scores_covid
# plots
vz_cpos_hist <- ggplot(vz_cpos, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "Virscan z-scores for covid+ samples, excluding outliers", x = "z-score (bin size 0.5)", y = "frequency") +
  theme_minimal()
# ggsave(filename = "vz_hist_3sd.png", plot = vz_hist_3sd, path = VIZ_PATH)

vz_cneg_hist <- ggplot(vz_cneg, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "Virscan z-scores for covid- samples, excluding outliers", x = "z-score (bin size 0.5)", y = "frequency") +
  theme_minimal()
# ggsave(filename = "vz_hist_3sd.png", plot = vz_hist_3sd, path = VIZ_PATH)


vz_cpos_hist
vz_cneg_hist
```


```{r}
# group by hospitalized vs. non-hospitalized 
# histograms based
vz_hosp <- pat_vz %>% filter(!is.na(Hospitalized == "hospitalized"))
vz_noth <- pat_vz %>% filter(!is.na(Hospitalized == "non-hospitalized"))

vz_hosp_summary <- vz_hosp %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

vz_noth_summary <- vz_noth %>%
  summarise(
    mean_z = mean(z_score, na.rm = TRUE),
    median_z = median(z_score, na.rm = TRUE),
    sd_z = sd(z_score, na.rm = TRUE),
    min_z = min(z_score, na.rm = TRUE),
    max_z = max(z_score, na.rm = TRUE),
    q1_z = quantile(z_score, 0.25, na.rm = TRUE),
    q3_z = quantile(z_score, 0.75, na.rm = TRUE),
    count = length(z_score)
  )

vz_hosp_summary
vz_noth_summary
```


```{r}
vz_hosp_hist <- ggplot(vz_hosp, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "Virscan z-scores for hospitalized samples, excluding outliers", x = "z-score (bin size 0.5)", y = "frequency") +
  theme_minimal()
# ggsave(filename = "vz_hosp_hist.png", plot = vz_hosp_hist, path = VIZ_PATH)

vz_noth_hist <- ggplot(vz_noth, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +  
  labs(title = "Virscan z-scores for non-hospitalized samples, excluding outliers", x = "z-score (bin size 0.5)", y = "frequency") +
  theme_minimal()
# ggsave(filename = "vz_noth_hist.png", plot = vz_noth_hist, path = VIZ_PATH)


vz_hosp_hist
vz_noth_hist
```




# 241105-TODOs
- plot stuff for hosp vs not hosp
- do logistic regression on cleaned data?
- random forest on cleaned data









```{r}

  
```




# TODO: remove; not so useful plots or intermediate steps 
histogram of original z-scores different resolution 
```{r}
# histogram of all z-scores
z_hist_log_binsize01 <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 0.1) +  # Adjust binwidth as needed
  labs(title = "histogram of log-scaled Virscan z-scores", x = "z-score (bin size = 0.1)", y = "log(frequency)") +
  xlim(-10, 10) +  # Adjust range to exclude extreme outliers
  theme_minimal()
# ggsave(filename = "pat_vz_hist_log_binsize01.png", plot = z_hist_binsize01, path = VIZ_PATH)

z_hist_log_binsize05 <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5) +  # Adjust binwidth as needed
  labs(title = "histogram of log-scaled Virscan z-scores", x = "z-score (bin size = 0.5)", y = "log(frequency)") +
  xlim(-10, 10) +  # Adjust range to exclude extreme outliers
  theme_minimal()
# ggsave(filename = "pat_vz_hist_log_binsize05.png", plot = z_hist_binsize05, path = VIZ_PATH)

z_hist_log_binsize01
z_hist_log_binsize05
```


```{r multiple-plots, fig.width=6, fig.height=4}
# histogram of all z-scores
z_hist_binsize01 <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 0.1) +  # Adjust binwidth as needed
  labs(title = "histogram of all Virscan z-scores", x = "z-score (bin size = 0.1)", y = "frequency") +
  xlim(-10, 10) +  # most values lie in this range
  theme_minimal()
ggsave(filename = "pat_vz_hist_binsize01.png", plot = z_hist_binsize01, path = VIZ_PATH)

z_hist_binsize05 <- ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5) +  # Adjust binwidth as needed
  labs(title = "histogram of all Virscan z-scores", x = "z-score (bin size = 0.5)", y = "frequency") +
  xlim(-10, 10) +  # Adjust range to exclude extreme outliers
  theme_minimal()
ggsave(filename = "pat_vz_hist_binsize05.png", plot = z_hist_binsize05, path = VIZ_PATH)

z_hist_binsize01
z_hist_binsize05
```


```{r}
z_hist_log_binsize01
z_hist_log_binsize05
```




```{r}
# histogram of all z-scores on log scale
 ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(binwidth = 1) +  # Adjust binwidth as needed
  labs(title = "histogram of all Virscan z-scores (log scale), including outliers", x = "z-Score", y = "log (frequency)") +
  scale_y_log10() +  
  theme_minimal()
 # ggsave(filename = "pat_vz_hist.png", path = VIZ_PATH)
```



```{r}
ggplot(pat_vz, aes(x = z_score)) +
  geom_histogram(aes(y = after_stat(count + 1)), binwidth = 0.5, fill = "gray") +  # Add offset to counts
  scale_y_log10(limits = c(1, 1e4)) +
  scale_x_continuous(limits = c(-10, 60)) +
  labs(title = "Histogram of All Virscan Z-Scores (Log Scale)", x = "Z-Score", y = "Log (Frequency)") +
  theme_minimal()

```



```{r}
z_hist_3sd_rs1e6 <- ggplot(pat_vz_rs, aes(x = z_score)) +
  geom_histogram(binwidth = 0.5, fill = "gray", color = "black") +
  labs(title = "histogram of 10^6 samples of Virscan z-scores", x = "z-score", y = "frequency") +
  theme_minimal()

z_hist_3sd_rs1e6

```

```{r}
z_hist_3sd_rs10e6_log <- ggplot(pat_vz_rs, aes(x = z_score)) +
  geom_histogram(aes(y = after_stat(count + 1)), binwidth = 0.5, fill = "gray", color = "black") +
  scale_y_log10() +
  labs(title = "histogram of 10^6 samples of Virscan z-scores, log-scaled", x = "z-score", y = "log(frequency)") +
  theme_minimal()

z_hist_3sd_rs10e6_log

```



