---
title: "Shrock2020_virscan_regression"
output: html_notebook
---


```{r}
library(Peptides)
library(readxl)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(data.table)
library(glmnet)
# set working directory to where script is
here::i_am("scripts/Shrock2020_virscan_regression.Rmd")
library(here)
source("../scripts/peptides.R")

```


```{r}
# load processed merged virscan data containing covid, hospitalization, peptide-zscores for each patient sample
# User supplies path to data directory, and file name they wish to read
DATA_DIR = "./data/processed"
PAT_META_VIR_Z_FILE_NAME = "pat_meta_vir_z_2024-10-22.rds"
pat_meta_vir_z <- readRDS(file = here::here(DATA_DIR, PAT_META_VIR_Z_FILE_NAME))
```

```{r}

# Convert to a tibble if needed
pat_meta_vir_z <- as_tibble(pat_meta_vir_z)

# Pivot to wide format using pivot_wider
pivot_pat_meta_vir_z <- pat_meta_vir_z %>%
  pivot_wider(names_from = id, 
              values_from = z_score)

pivot_pat_meta_vir_z <- pivot_pat_meta_vir_z %>% mutate(COVID19_status = if_else(COVID19_status == "positive", 1, 0),
                                  Hospitalized = if_else(Hospitalized == "hospitalized", 1, 0))
                                  

# View the resulting pivoted data
head(pivot_pat_meta_vir_z, 20)
```

```{r}
# # save pat_meta_vir_z; ready to use as input to model
# saveRDS(pivot_pat_meta_vir_z, "../data/processed/pat_meta_vir_z_input.rds")
```


```{r}

# response: covariates: all z scores
NUM_PEPTIDES = 115753
covstat_zscores_str <- paste("COVID19_status~", paste(1:NUM_PEPTIDES, collapse="+"), sep = "")
# lm_z_means_feats_sc <- lm(lm_str, data = vir_merged)
# summary(lm_z_means_feats_sc)
# save_as_rds(lm_z_means_feats_sc, "lm_z_means_feats_sc", "./results/")

## Try LR just because, try  LASSO, Elastic Net
## Otherwise, want to try dim reduction first 


set.seed(1031)
## Fit LASSO
## find penalty \lambda by cross-validation by minimizing MSE
z_scores <- as.matrix(pivot_pat_meta_vir_z[, -c(1:5)]) # just the z-score columns
cov_stat <- pivot_pat_meta_vir_z$COVID19_status

## filter patients for whom we have hospitalization status
pat_vir_h <- pivot_pat_meta_vir_z %>% filter(!is.na(Hospitalized))
z_scores_h <- as.matrix(pat_vir_h[, -c(1:5)]) # just the z-score columns
hos_stat <- pat_vir_h$Hospitalized

lasso_cov_z <- cv.glmnet(x=z_scores, y=cov_stat, alpha=1,family="binomial",type.measure = "mse")
save_as_rds(lasso_cov_z, "lasso_cov_z_vir", "./results")

lasso_hos_z <- cv.glmnet(x=z_scores_h, y=hos_stat, alpha=1,family="binomial",type.measure = "mse")
save_as_rds(lasso_hos_z, "lasso_hos_z_vir", "./results")

## Fit Ridge
ridge_cov_z <- cv.glmnet(x=z_scores, y=cov_stat, alpha=0,family="binomial",type.measure = "mse")
save_as_rds(ridge_cov_z, "ridge_cov_z_vir", "./results")

ridge_hos_z <- cv.glmnet(x=z_scores_h, y=hos_stat, alpha=0,family="binomial",type.measure = "mse")
save_as_rds(ridge_hos_z, "ridge_hos_z_vir", "./results")

```

```{r,  figures-side, fig.show="hold", out.width="50%"}
plot(lasso_hos_z)
title("LASSO hospitalization status vs. all z-scores")

plot(lasso_cov_z)
title(main="LASSO covid status vs. all z-scores")

```

```{r}
plot(ridge_cov_z)
title(main="Ridge covid status vs. all z-scores")
```

```{r}
l_min <- lasso_cov_z$lambda.min
coef <- coef(lasso_cov_z, s = "lambda.min")
# should give the same results as using coef() on cv.glmnetobject and specifying s
# lasso_m_cov_z <- glmnet(x = z_scores, y = cov_stat, alpha = 1, lambda = l_min)
# coef_2 <- lasso_m_cov_z$beta
nonzero_coef <- coef[coef[,1]!=0,][-1] # also exclude intercept

l_h_min <- lasso_hos_z$lambda.min
coef_h <- coef(lasso_hos_z, s = "lambda.min")
nonzero_coef_h <- coef_h[coef_h[,1]!=0,][-1]

l_r_min <- ridge_cov_z$lambda.min
coef_r <- coef(ridge_cov_z, s = "lambda.min")
nonzero_coef_r <- coef_r[coef_r[,1]!=0,][-1]



```


```{r}

## Relating LASSO coefficients to peptides
vir_lib <- readRDS(file = here::here("./data/processed", "vir_lib_2024-10-24.rds"))
cov_lib <- readRDS(file = here::here("./data/processed", "cov_lib_2024-10-24.rds"))

nonzero_coef_df <- data.frame(
  id = names(nonzero_coef), # extract peptide ids
  coef = nonzero_coef, # extract nonzero coefficients
  stringsAsFactors = FALSE
)

nonzero_coef_h_df <- data.frame(
  id = names(nonzero_coef_h), # extract peptide ids
  coef = nonzero_coef_h, # extract nonzero coefficients
  stringsAsFactors = FALSE
) %>% mutate(id = as.numeric(id))


## Relate Ridge coefficients to peptides 
nonzero_coef_r_df <- data.frame(
  id = names(nonzero_coef_r), # extract peptide ids
  coef = nonzero_coef_r, # extract nonzero coefficients
  stringsAsFactors = FALSE
) %>% mutate(id = as.numeric(id))

# make sure id values in nonzero_coef_df and vir_lib match (want both to be numeric)
nonzero_coef_df <- nonzero_coef_df %>% 
  mutate(id = as.numeric(id))

# map back to peptides by descending order of LASSO coefficients
tmp_desc_coef <- vir_lib %>% 
  inner_join(nonzero_coef_df, by = "id") %>%
  select(coef, everything()) %>%
  arrange(desc(coef))

tmp_desc_coef_h <- vir_lib %>% 
  inner_join(nonzero_coef_h_df, by = "id") %>%
  select(coef, everything()) %>%
  arrange(desc(coef))


tmp_desc_coef_r <- vir_lib %>% 
  inner_join(nonzero_coef_r_df, by = "id") %>%
  select(coef, everything()) %>%
  arrange(desc(coef))

```

## Top 20 peptides in Virscan library based on LASSO coefficients (response: covid, binary)

```{r}
head(tmp_desc_coef[c("coef", "Organism", "Species")], 50)
```

```{r}





```

## Top 20 peptides in Virscan library based on LASSO coefficients (respose: hospitalization, binary)

```{r}
head(tmp_desc_coef_h[c("coef", "Organism", "Species")], 50)
```


## Top 20 peptides in Virscan library based on Ridge coefficients (response: covid, binary)

```{r}
head(tmp_desc_coef_r[c("coef", "Organism", "Species")], 20)
```

## TODO
## Top 20 peptides in Virscan library based on Ridge coefficients (response: hospitalization, binary)

```{r}
head(tmp_desc_coef_h_r[c("coef", "Organism", "Species")], 20)
```

